---
title: "Introduction to usgeogr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{usgeogr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE}
library(usgeogr)
```

Economists often utilize variation across geographic regions as natural
experiments. For example, GHKS (2019) utilize differences in the minimum wage
across state borders over time to estimate its employment effects.[^1]

[^1]: Gopalan, Hamilton, Kalda, and Sovich (2019). "State minimum wage changes and employment:
Evidence from one million hourly wage workers". Working paper.

The `usgeogr` package helps streamline such analyses by providing a set of datsets and functions 
for commonly studied U.S. geographic units that can be quickly combined with spatial identification
strategies. 

This document introduces you to `usgeogr`'s datasets and functions. For more information,
please see the [Github](https://github.com/davidsovich/usgeogr). 

## State data

`state_df` provides a list of all U.S. states and their postal codes.

## County data

### Baseline datasets

`county_df` provides a list of all counties in the continental United States. The fields include 
each county's FIPS identifier, population (as of 2010), and the latitude and longitude
of its population center. 

`adjacent_county_df` provides a list of all adjacent counties in the continental United States.
Each row corresponds to a county and its adjacent neighbor. Each adjacent county pair will appear
twice in the dataset. 

### Cross-border datasets for county-level studies

`cbcp_df` provides a list of all adjacent cross-border county pairs in the United States.[^2] Each row 
corresponds to a border county. If a border county has $p >1$ cross-border neighbors, then it will
have $p>1$ entries in the dataset. There are more unique cross-border county pairs 
(1,308) than unique border counties (1,184). 

[^2]: An adjacent cross-border county pair is defined as a pair of adjacent border counties that
reside in different states. A single border county can belong to multiple adjacent cross-border 
county pairs. The definition can be traced to Dube, Lester, and Reich (ReSTAT, 2010).

### Cross-border datasets for within county-level studies

Identification strategies that exploit variation within cross-border county pairs cannot be easily 
transported to analyses in which the unit of analysis is more granular than the county-level (e.g.,
the individual level). In such cases, each border county must be assigned to a distinct "cluster"
for which to exploit within-cluster variation across units. Unfortunately, there is no unique 
method for constructing such "clusters".[^3] To address the problem of non-uniqueness, this packages
provides four possible mappings: 

1. `sbscp_df` uniquely maps each border county to an adjacent state border pair segment (e.g., 
AL-FL). If a border county resides on multiple state border pair segments, then the 
border segment closest to the county's population center is retained. A small set of counties are
mapped to a state border pair segment with only one state or county represented.[^4]

2. `cpcp_df` uniquely maps each border county to a "couplet" of two counties. Uses an algorithm
that maximizes the number of distinct clusters with exactly two cross-border members. Because of 
the strict requirement of two members per cluster, a substantial fraction of the counties in the
data are left unmatched. 

3. `rcpcp_df` relaxes the strict membership cap in the `cpcp_df` and maps unmatched counties to
their closest cross-border neighbor pair. 

4. `mxcp_df` assings each county to a grouping formed by its bordering county (including itself)
with the most direct cross-border connections. 

5. `rmxcp_df` places unmathced into BLAH. 

[^3]: This is because some counties have several cross-border adjacent neighbors. For example, 
counties may be geographically staggered so that each county borders approximately 2 other 
cross-state counties (creating overlap). Only if each cross-broder county had a single cross-border
neighbor could we have a unique and natural mapping of pairs for the groupings. However, there is
not a single such unique mapping in the United States.

[^4]: Hence, variation from these counties is discarded when conducting comparisons across 
state borders within state border segment pairs. 

## ZIP code data

### Baseline datasets

`zip_df` provides a list of all ZIP codes in the continental United States. Information on 
population, centroid location, and land usage is populated for ZIP codes that map into ZCTAs.

### Cross-border datasets for ZIP code and within-ZIP code studies

BLAH

## Functions

### County functions

BLAH

### ZIP code functions

BLAH













